---
title: "Kolmogorov-Smirnov-test.R"
author: "Vinzent Aschir, Balaji Govindarajan and Ankit Kummar"
date: "2025-11-08"
output:
  pdf_document:
    latex_engine: xelatex

---

# empirical_CDF_and_Kolmogorov_Test

> **Instruction:** Study the empirical cumulative distribution function (ECDF) and the Kolmogorov–Smirnov test.Test whether a dataset follows a given distribution, or whether two samples come from the same one. Visualize ECDFs (implementated by yourself) and interpret.

for following the project on your own clone by:

```bash
git clone https://github.com/Aschir/empirical_CDF_and_Kolmogorov_Test/
```

## theory about Kolmogorov-Smirnov test (KS-Test)

### One-sample Kolmogorov-Smirnov Test

The one-sample Kolmogorov–Smirnov (KS) test quantifies the difference between the empirical distribution of a dataset and a theoretical distribution and can be applied when your analysed data is  iid. 

### Example:

Testing if \$\$X \\sim N(\\mu, \\sigma\^2) \$\$

A basketball player Mike, shoots each training 1000 times. he has 200 trainings and scores with a probability of p=0.5, does his score follow a normal distribution?

$$ training \sim Binom(n=1000, p=0.5) $$
with 100 performed trainings: 
```{r}
set.seed(42)
# data 
n = 1000 
prob = 0.5 
n_training = 100 

X = rbinom(n_training, size = n, prob = prob)
```

$$H_0: X\sim N(\mu = 500, \sigma^2)$$

$$H_1: X \not\sim N(\mu = 500, \sigma^2)$$

Let's have a look in it's distribution:

```{r}
set.seed(42)

# for visualisation:  10% quantile and closest point to 10% quantile
q10 <- quantile(X, 0.1)                      
closest_q10 <- X[which.min(abs(X - q10))]   

# ---- Plot 1: the distribution of X ----
plot(1:length(X), X,
     xlab = "training index (sample)", 
     ylab = "Scores in this training",
     main = "Distribution of scores in the progress of trainings.",
     pch = 16, 
     col="grey")

# ---- Plot 2: Histogram + part shaded + theoretical curve ----
h = hist(X,
         breaks = 50,
         probability = TRUE,
         main = paste0("Distribution of scored points"),
         xlab = "Scores in a training",
         col = "lightgray",
         border = "white")

# --- Shade bars up to closest 10% quantile ---
for (i in 1:(length(h$breaks)-1)) {
  left  <- h$breaks[i]
  right <- h$breaks[i+1]
  mid   <- (left + right) / 2
  
  if (mid <= closest_q10) {
    rect(left, 0, right, h$density[i],
         col = rgb(1, 0, 0, 0.4),
         border = NA)
  }
}

# --- Theoretical distribution ---
x_vals <- seq(min(X), max(X), length.out = 200)
theo_density <- dnorm(x_vals, mean = 500, sd = sqrt(250))

lines(x_vals, theo_density, col = "blue", lwd = 3)

# --- Legend ---
legend(
  "topright",
  legend = c(paste0("P(X ≤ ", closest_q10, ")"), 
             "theo. distr. (H0)", 
             "distr. data"),
  pch    = 22,                          
  pt.cex = 2,
  pt.bg  = c(rgb(1,0,0,0.4), "blue", "lightgray"),  # fill colors
  col    = c(rgb(1,0,0,0.4), "blue", "lightgray"),  # border colors
  bty    = "n"
)
```

In a general setting,

Null hypothesis: $H_0: F_X = F_0$\
Alternative hypothesis: $H_1: F_X \neq F_0$

and our test statistic is:

$$
D_n = \sup_x |F_n(x) - F(x)|
$$

We make no assumptions on the parameter of the given distribution! Only the CDF 

We Compute the CDF:

```{r}
set.seed(42)

#CDF (ECDF) of the given DATA. 
  ecdf_X = ecdf(X) # we use the given func for the data 
  
  plot(ecdf_X,
       main = "Empirical CDF of X",
       xlab = "x",
       ylab = "F(x)",
       verticals = TRUE,
       do.points = FALSE)

# theoretical cdf in manual computation of the cdf 
  x_vals = seq(min(X), max(X), length.out = 200)
  cdf_theo <- pnorm(x_vals, mean = 500, sd = sqrt(250))

  lines(x_vals, cdf_theo, col = "blue", lwd = 2)

# Add the point seen in the previous data 
  y_q10 <- ecdf_X(closest_q10)
  
  points(closest_q10, y_q10, col = rgb(1, 0, 0, 0.4), pch = 16)

# ---- Add vertical line for sup difference (Kolmogorov D) ----

  # max difference computation
  emp_vals <- ecdf_X(x_vals)
  diffs <- abs(emp_vals - cdf_theo)
  D <- max(diffs)
  
  
  x_value_f <- x_vals[which.max(diffs)]
  y_emp_at_f <- emp_vals[which.max(diffs)]
  y_theo_at_f <- cdf_theo[which.max(diffs)]
  eps <- 1e-8

  # draw the vertical line of max difference
  #segments(x0 = x_value_f, y0 = y_emp_at_f,
  #       x1 = x_value_f, y1 = y_theo_at_f,
  #       col = "red", lwd = 2, lty = 2)
  


  # mark max difference points in their distributions
  points(x_value_f, y_emp_at_f, col="red", pch=16)
  points(x_value_f, y_theo_at_f, col="blue", pch=16)
  lines(x = c(x_value_f-eps, x_value_f+eps), y = c(y_emp_at_f, y_theo_at_f),
      col = "red", lwd = 2, lty = 2)
  
  legend("topleft",
       legend = c("ECDF of X", 
                  "Theoretical CDF", 
                  paste0("Max difference D = ", round(D, 3)),
                  paste0("P(X ≤ ", closest_q10, ")")),
       col = c("black", "blue", "red", rgb(1,0,0,0.4)),
       lwd = c(1, 1, 1, NA),
       lty = c(1, 1, 1, NA),
       pch = c(NA, NA, 1, 16),
       bty = "n")

```

we used the maximum as a numerical approximation of the supremum.

the supremum would consider all points not only the ones in the dataset.

the maxium deviation difference is not normal distributed but follows a Kolmogorov-Smirnov deviations- distribution defined by:

The maximum deviation $D_n$ is **not normally distributed**. Instead, it follows the **Kolmogorov distribution**, which has the cumulative distribution function:

$$
P(\sqrt{n} D_n \le \lambda) = 1 - 2 \sum_{k=1}^{\infty} (-1)^{k-1} e^{-2 k^2 \lambda^2}
$$

which is in r:

```{r}
ks_pvalue <- function(D, n, maxK=1000){
  lambda <- sqrt(n) * D
  k <- 1:maxK
  pval <- 2 * sum((-1)^(k-1) * exp(-2 * k^2 * lambda^2))
  pval <- min(max(pval, 0), 1)
  return(pval)
}
```

now we compute the p-value: and form our decision boundary:

\\[\\text{p-value} = P(D_n \\ge D\_{\\text{obs}} \\mid H_0)\\]

```{r}
p_value <- ks_pvalue(D, length(X))
p_value

# Decision based on alpha
alpha <- 0.05
if(p_value < alpha){
  cat("Reject H0: the data does not follow the theoretical distribution.\n")
} else {
  cat("Fail to reject H0: no evidence against the theoretical distribution.\n")
}
```

we can compare our computation with the builtin-method:

the results are not very accurate since we used the maximum instead of the minimum and for sakes of explainability an discrete distribution function instead of an continious function

```{r}
# Built-in KS test
ks_result <- ks.test(X, "pnorm", mean = 500, sd = sqrt(250))
ks_result

```