---
title: "MSc Data Science: R for Data Science"
author: "PhD Student Raphaël Razafindralambo (Email: raphael.razafindralambo@inria.fr)"
date: "Last updated on `r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Scripts

> RStudio is the preferred integrated development environment (IDE) for writing and running R scripts. Installation steps are detailed in the notebook of the first session.

An **R Script** is a simple text file (usually with the `.R` extension) that contains a sequence of R commands.\
It allows you to **save**, **reuse**, and **share** your code easily, instead of typing commands directly into the console.

You can create an R Script in RStudio by clicking:\
**File → New File → R Script**\
and then run lines of code with:\
**Ctrl + Enter** (Windows) or **Cmd + Enter** (Mac).

## Rmarkdown

Rmarkdown is a comprehensive way of creating reports which involve R (or other languages). A synthetic presentation of the Rmarkdown system is available here: <https://rmarkdown.rstudio.com/lesson-1.html>. For advanced details on the language, please refer to <https://bookdown.org/yihui/rmarkdown/>.

The basic commands of Rmarkdown:

-   the `#` allows to organize the document in sections, subsections, ... The more you add `#`, the more you go down into the subsections, subsubsections, ...

-   the `*` allows to higlight some part of the text: if you use one `*tyty*` it produces *tyty*. If you use `**tyty**`, the text **tyty** will be in bold.

-   the `>` allows to create some "citation / remark box": \> The citation that I would like to highlight!

-   It is possible to provide online code using single quote. As an exemple: `ls()`.

It is also possible to provide parts of code, that are enventually run:

```{r}
rnorm(15)
```

There are several options:

-   the first one is the language to use (here, R),
-   the `echo` option allows to disable the display of the output. Notice that the code is however run and you can use later its ouptput!

```{r, echo=FALSE}
mu = 2; sigma = 1
x = rnorm(1500,mu,sigma)
hist(x,breaks = 50,freq = FALSE, col='lavender')
lines(density(x),col='red',lwd=2)
```

-   the `include` option allows to hide both the code and the output (even though the code is run and the output exists in the memory).
-   the `eval` option disables the execution of the code.

```{r, include=FALSE, eval=FALSE}
if(!require(beepr)) {
  install.packages("beepr", repos = "http://cran.us.r-project.org")
  library(beepr)
}

```

The different outputs for the Rmd file are either HTML, PDF or Word documents.

## Shiny

The `shiny` package for R is a great way to creat interfaces with some R codes. A shiny code creates a web page with JavaScript which allows the user to execute some (hidden) R code and look at the results in a very easy way.

```{r, eval=FALSE}
library(shiny)
runExample("01_hello")
```

To start a Shiny app, you have to create two files:

-   `ui.R` is devoted to design the user interface,
-   `server.R` is running the R codes conditionnaly to the options set by the user through the interface.

The two minimal files [ui.R](./ui.R) and [server.R](./server.R) can be used to start a new app project. For more details on the desiging of Shiny App, please refer to <http://shiny.rstudio.com/tutorial/>.

------------------------------------------------------------------------

## Projects

This project aims to **explore and understand real-world applications of R** through small, hands-on experiments on statistical methods and data analysis.\
You will learn not only to *implement* statistical tools, but also to *explain them pedagogically* to others.

The focus is both **technical** and **educational**:\
you should be able to demonstrate that you understand *how* the method works, *how to implement it in R*, and *how to explain it clearly*.

### Objectives

-   Learn to implement a method in R, **both manually and using built-in packages**.\
-   Apply your implementation to **"real" datasets** if possible.\
-   Understand and **explain step by step** how the method works in R.\
-   Prepare a short, clear, and pedagogical presentation of your results.

### Monitoring sessions

There will be **three mandatory sessions** dedicated to the project, during which you will work in groups and receive guidance. This project is **not extensive**; it is designed to be completed during the three sessions.

### Final Presentation

The presentations will take place **during the week of December 8**.Each group will have a short time slot to present their work in front of the class.

You are **free to choose the presentation format**, as long as it is **clear and easy to follow**. For example, you may present directly from your **RMarkdown**, use a few **slides (PowerPoint, PDF)**, or simply **show your analysis interactively with a notebook**.

Each presentation will last **10 minutes**, followed by **5 minutes of questions**. Questions will be asked by **myself** and by **one other group chosen at random**

### Evaluation

Each group will receive **two grades**:

| Component | Description | Weight |
|----|----|----|
| Presentation | Clarity, pedagogy, correctness of implementation, and quality of explanations | 70% |
| Questions | Quality and relevance of the question asked to another group | 30% |

*Note: The grading scheme may be adjusted if necessary.*

### List of projects

Each group will be composed of **two students**. Every group must choose one project topic, and no two groups may work on the same project. The assignment rule is *first come, first served*.

A Google Sheet will be shared for you to fill in your **group members** and **chosen project**. You will have **until November 12 (!!)** to form your group and select your topic.

#### 1. Statistical Tests of Independence

Explore how to test whether two categorical variables are statistically independent.\
Use methods such as the **Chi-squared test** (between others) and interpret the results.

------------------------------------------------------------------------

#### 2. Statistical Student Tests

Compare the means of one or two groups using **Student’s t-tests**. Implement the t-statistic manually and compare results with the `t.test()` function in R.

------------------------------------------------------------------------

#### 3. Statistical Tests for Benchmarking (McNemar’s Test)

Use **McNemar’s test** to compare the performance of two classifiers on the same dataset. It is particularly useful in **machine learning** for comparing confusion matrices on paired binary outcomes.

------------------------------------------------------------------------

#### 4. Empirical CDF and Kolmogorov Tests

Study the **empirical cumulative distribution function (ECDF)** and the **Kolmogorov–Smirnov test**.\
Test whether a dataset follows a given distribution, or whether two samples come from the same one. Visualize ECDFs (implementated by yourself) and interpret.

------------------------------------------------------------------------

#### 5. K-Means Clustering

Implement the **k-means algorithm** from scratch and compare it with R’s `kmeans()` function.\
Apply it to a dataset to discover natural clusters and visualize the partitioning.\
Discuss initialization effects and convergence. It is also worth exploring the K-means++ initialization.

------------------------------------------------------------------------

#### 6. Principal Component Analysis (PCA)

Perform and analyse **dimensionality reduction** using PCA.\
Implement PCA step by step using covariance matrices and eigenvalues and reproduce the same analysis with a standard R package.

------------------------------------------------------------------------

#### 7. Linear Regression

Study **linear regression** models on real-world datasets.\
You may use `lm()` to train your model. Use diagnostic plots to assess model fit and residual assumptions.

------------------------------------------------------------------------

#### 8. Deep Learning in R with torch

Explore the basics of **deep learning** using the `torch` package in R.\
Implement a simple neural network for regression or classification.\
Experiment with training, loss functions, and visualizing learning curves.

------------------------------------------------------------------------

#### 9. Bootstrap Method (Confidence Intervals)

Use **bootstrap resampling** to estimate confidence intervals for means, medians, regression coefficients, ... Implement the procedure manually and compare with R’s `boot` package.\
Interpret variability and bias through simulations.

------------------------------------------------------------------------

#### 10. Random Forests

Implement and analyze **Random Forests** for classification or regression using the associated R package.\
Study the concept of bagging and variable importance.\
Optionally, attempt a simple manual implementation of the averaging principle.
