---
title: "Kolmogorov–Smirnov-test.R"
author: "Vinzent Aschir, Balaji Govindarajan and Ankit Kummar"
date: "2025-11-08"
output: html_document
---

# empirical_CDF_and_Kolmogorov_Test

Instruction: Study the empirical cumulative distribution function (ECDF) and the Kolmogorov–Smirnov test.  
Test whether a dataset follows a given distribution, or whether two samples come from the same one.  
Visualize ECDFs (implemented by yourself) and interpret.

For following the project on your own clone by:

`git clone https://github.com/Aschir/empirical_CDF_and_Kolmogorov_Test/`

---

## Theory about Kolmogorov–Smirnov test (KS-Test)

### One-sample Kolmogorov–Smirnov Test

The one-sample Kolmogorov–Smirnov (KS) test quantifies the difference between the empirical distribution
of a dataset and a theoretical distribution and can be applied when your analysed data is iid.

Example: Testing if \( X \sim N(\mu, \sigma^2) \)

A basketball player Mike shoots each training 1000 times. He has 200 trainings and scores with a probability
of \( p = 0.5 \). Does his score follow a normal distribution?

\[
\text{training} \sim \text{Binom}(n = 1000, p = 0.5)
\]

with 100 performed trainings:

```{r}
set.seed(42)
# data
n = 1000
prob = 0.5
n_training = 100
X = rbinom(n_training, size = n, prob = prob)
```

Hypotheses:

\[
H_0 : X \sim N(\mu = 500, \sigma^2) \\
H_1 : X \nsim N(\mu = 500, \sigma^2)
\]

Let’s have a look at its distribution:

```{r}
set.seed(42)
# for visualisation: 10% quantile and closest point to 10% quantile
q10 <- quantile(X, 0.1)
closest_q10 <- X[which.min(abs(X - q10))]

# ---- Plot 1: the distribution of X ----
plot(1:length(X), X,
     xlab = "training index (sample)",
     ylab = "Scores in this training",
     main = "Distribution of scores in the progress of trainings.",
     pch = 16,
     col ="grey")
```

```{r}
# ---- Plot 2: Histogram + part shaded + theoretical curve ----
h = hist(X,
         breaks = 50,
         probability = TRUE,
         main = "Distribution of scored points",
         xlab = "Scores in a training",
         col = "lightgray",
         border = "white")

# --- Shade bars up to closest 10% quantile ---
for (i in 1:(length(h$breaks) - 1)) {
  left <- h$breaks[i]
  right <- h$breaks[i+1]
  mid <- (left + right) / 2
  if (mid <= closest_q10) {
    rect(left, 0, right, h$density[i],
         col = rgb(1, 0, 0, 0.4),
         border = NA)
  }
}

# --- Theoretical distribution ---
x_vals <- seq(min(X), max(X), length.out = 200)
theo_density <- dnorm(x_vals, mean = 500, sd = sqrt(250))
lines(x_vals, theo_density, col = "blue", lwd = 3)

# --- Legend ---
legend("topright",
       legend = c(paste0("P(X ≤ ", closest_q10, ")"),
                  "theo. distr. (H0)",
                  "distr. data"),
       pch = 22,
       pt.cex = 2,
       pt.bg = c(rgb(1,0,0,0.4), "blue", "lightgray"),
       col = c(rgb(1,0,0,0.4), "blue", "lightgray"),
       bty = "n")
```

In a general setting,

- Null hypothesis: \( H_0 : F_X = F_0 \)  
- Alternative hypothesis: \( H_1 : F_X \ne F_0 \)

and our test statistic is:

\[
D_n = \sup_x \big| F_n(x) - F(x) \big|
\]

We make no assumptions on the parameter of the given distribution! Only the CDF.

We compute the CDF:

```{r}
set.seed(42)
# CDF (ECDF) of the given DATA.
ecdf_X = ecdf(X) # we use the given func for the data
plot(ecdf_X,
     main = "Empirical CDF of X",
     xlab = "x",
     ylab = "F(x)",
     verticals = TRUE,
     do.points = FALSE)

# theoretical cdf in manual computation of the cdf
x_vals = seq(min(X), max(X), length.out = 200)
cdf_theo <- pnorm(x_vals, mean = 500, sd = sqrt(250))
lines(x_vals, cdf_theo, col = "blue", lwd = 2)

# Add the point seen in the previous data
y_q10 <- ecdf_X(closest_q10)
points(closest_q10, y_q10, col = rgb(1, 0, 0, 0.4), pch = 16)

# max difference computation
emp_vals <- ecdf_X(x_vals)
diffs <- abs(emp_vals - cdf_theo)
D <- max(diffs)
x_value_f <- x_vals[which.max(diffs)]
y_emp_at_f <- emp_vals[which.max(diffs)]
y_theo_at_f <- cdf_theo[which.max(diffs)]
eps <- 1e-8

points(x_value_f, y_emp_at_f, col="red", pch=16)
points(x_value_f, y_theo_at_f, col="blue", pch=16)
lines(x = c(x_value_f-eps, x_value_f+eps), y = c(y_emp_at_f, y_theo_at_f),
      col = "red", lwd = 2, lty = 2)

legend("topleft",
       legend = c("ECDF of X",
                  "Theoretical CDF",
                  paste0("Max difference D = ", round(D, 3)),
                  paste0("P(X ≤ ", closest_q10, ")")),
       col = c("black", "blue", "red", rgb(1,0,0,0.4)),
       lwd = c(1, 1, 1, NA),
       lty = c(1, 1, 1, NA),
       pch = c(NA, NA, 1, 16),
       bty = "n")
```

We used the maximum as a numerical approximation of the supremum.
The supremum would consider all points, not only the ones in the dataset.

The maximum deviation difference is not normally distributed but follows a Kolmogorov–Smirnov deviation distribution defined by:

The maximum deviation \(D_n\) is not normally distributed. Instead, it follows the Kolmogorov distribution, which has the cumulative distribution function

\[
P(\sqrt{n} D_n \le \lambda) = 1 - 2 \sum_{k=1}^{\infty} (-1)^{k-1} e^{-2 k^2 \lambda^2}
\]

which is in R:

```{r}
ks_pvalue <- function(D, n, maxK = 1000){
  lambda <- sqrt(n) * D
  k <- 1:maxK
  pval <- 2 * sum((-1)^(k-1) * exp(-2 * k^2 * lambda^2))
  pval <- min(max(pval, 0), 1)
  return(pval)
}

# compute p-value
p_value <- ks_pvalue(D, length(X))
p_value

# Decision based on alpha
alpha <- 0.05
if (p_value < alpha) {
  cat("Reject H0: the data does not follow the theoretical distribution.\n")
} else {
  cat("Fail to reject H0: no evidence against the theoretical distribution.\n")
}
```

We can compare our computation with the builtin method:

```{r}
# Built-in KS test
ks_result <- ks.test(X, "pnorm", mean = 500, sd = sqrt(250))
ks_result
```

The results are not very accurate since we used the maximum instead of the minimum and, for sake of explainability, a discrete distribution function instead of a continuous function.

---

## Project Continuity: Theory vs. Application (The K-S Test)

Also, we will run both the One-Sample and the Two-Sample K-S Tests on the practical geyser 
data to prove it is not a single bell curve, but is actually made of two separate and distinct 
schedules.

| Parameter            | Theoretical Component                                                                 | Practical Component                                                                 |
|----------------------|---------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------|
| Data Source          | Used an artificially generated dataset (Binomial simulation) to ensure clean, controlled data for formula testing. | Will use a real-world dataset (faithful geyser waiting times) known to be messy and bimodal, representing a true challenge. |
| Primary Goal         | Confirmed a hypothesis (failed to reject the Null Hypothesis), proving the K-S test correctly identifies a clean, theoretical distribution. | Will challenge a hypothesis (will reject the Null Hypothesis), proving that the real data is too complex for simple models and requires deeper analysis. |
| Core Method          | Focused on formula validation by manually calculating the K-S p-value.               | Will include the manual ECDF plot and the two distinct distribution plot.          |
| ECDF Visualization   | Visualized the ECDF of the observed data against the theoretical distribution to check for fit. | Will manually create the ECDF step chart to prove foundational understanding of how the data accumulates. |

---

## Project Goals – For this messy, bimodal, real world dataset

- Verifying Distribution: We use statistics to formally prove that the geyser data, which visibly shows two different schedules (short and long waits), is not a simple, single "bell curve."
- Manual ECDF Implementation: We manually build the cumulative step chart (ECDF) to show exactly how the data accumulates from 0% to 100%.
- One-Sample K-S Test: We run this test to reject the bell curve assumption, stopping us from making the critical mistake of using a simple, wrong forecasting model.
- Two-Sample K-S Test: We split the data into two groups (Short Waits and Long Waits) and use this test to prove that they are drawn from two separate and distinct distributions.

### Advantages and Applications (Why It Matters)

- Prevents Model Failure: The most critical advantage is preventing the misapplication of statistical models. Ignoring the bimodal nature of the data would lead to unreliable forecasting results if a simple Normal model were used.

---

# Project 4: Empirical CDF and Kolmogorov Tests – Steps

### Step 1: Set Up the Environment

Started coding on blank R Script in RStudio.

### Step 2: Load Data and Library

In this step, we load the `ggplot2` plotting tool and the `faithful` dataset. After quickly inspecting the data structure (`str`, `head`), we extract the vital `waiting` column and save it specifically as `waiting_times`. The step concludes by running a summary to establish the initial statistics for this new variable.

- `str(faithful)`: This shows you the names of the variables and their types.  
- `head(faithful)`: View the first few rows to understand the data.

Basically, inspecting the structure of the dataset, then getting descriptive statistics of the variable `waiting_times`.

```{r setup-faithful, message=FALSE}
library(ggplot2)
data(faithful)
str(faithful)
head(faithful)
waiting_times <- faithful$waiting
summary(waiting_times)
```

### Step 3: Inspect Data

Plotted a histogram on `waiting_times` to visually confirm its bimodal nature, which indicates it is not a Normal distribution.

```{r hist-waiting}
hist(waiting_times,
     main = "Distribution of Geyser Waiting Times",
     xlab = "Waiting Time (minutes)",
     col = "lightblue",
     border = "white")
```

Why This Makes the Project "Interesting":

- Because the histogram shows two humps, it tells us that the geyser is not operating on one single, random schedule. It strongly suggests there are two different natural states: a "short cycle" (around 55 minutes) and a "long cycle" (around 80 minutes).
- Our project's entire purpose is to take that visual suggestion from the histogram and use advanced statistical tools (the ECDF and K-S tests) to mathematically prove that those two schedules are real and separate.

### Step 4: Sort the Data (ECDF Step 1)

Sort the data points, which will serve as the x-values for the ECDF plot.

```{r ecdf-step1}
data_sorted <- sort(waiting_times)
```

### Step 5: Calculate Proportions (ECDF Step 2)

Determine the total number of observations (n) and calculate the cumulative proportion (i/n) for each sorted point.

```{r ecdf-step2}
n <- length(data_sorted)
ecdf_values <- (1:n) / n
```

### Step 6: Prepare Plotting Data

Combine the sorted data and the proportions into a single data frame for visualization.

```{r ecdf-step3}
ecdf_df_manual <- data.frame(
  x = data_sorted,
  Fn = ecdf_values
)
```

### Step 7: Visualize the ECDF

Create a plot using ggplot2 and `geom_step()` to generate the required step-function visualization of the manually calculated ECDF.

```{r ecdf-plot}
ggplot(ecdf_df_manual, aes(x = x, y = Fn)) +
  geom_step(color = "darkred", size = 1) +
  labs(
    title = "Manually Implemented Empirical CDF",
    x = "Waiting Time (minutes)",
    y = "F_n(x)"
  )
```

The chart is mathematically perfect because it correctly plots the cumulative percentage from 0% to 100% using steps, and clearly shows the two distinct clusters where the curve rises steeply.

### Step 8: Estimate Normal Parameters

Calculate the mean (mu) and standard deviation (sigma) of the data, as these are required parameters for the theoretical Normal distribution in the K-S test.

```{r normal-params}
mu <- mean(waiting_times)
sigma <- sd(waiting_times)
mu
sigma
```

### Step 9: Perform One-Sample K-S Test

Execute the test against the theoretical Normal CDF (`pnorm`) defined by the calculated mu and sigma.

```{r one-sample-ks-faithful}
ks_test_normal <- ks.test(waiting_times, "pnorm", mean = mu, sd = sigma)
ks_test_normal
```

### Step 10: Interpret K-S Test 1

The results from the one-sample Kolmogorov–Smirnov test show a test statistic of **D = 0.15536** and a **p-value of 3.967e-06**.

Since the p-value (0.000003967) is far below the conventional significance level of \( \alpha = 0.05 \), we must **reject the Null Hypothesis (H_0)** that the data is drawn from a Normal distribution defined by the sample's mean and standard deviation.

This statistical rejection strongly validates our initial visual assessment of the histogram, which showed the data is clearly **bimodal** (has two peaks), making it fundamentally incompatible with the single-peaked, bell-shaped characteristics of the Normal distribution.

### Step 11: Create Two Samples

Split the data into two distinct groups based on the median to prepare for the two-sample K-S test.

```{r split-samples}
median_time <- median(waiting_times)
sample_A <- waiting_times[waiting_times <= median_time]
sample_B <- waiting_times[waiting_times > median_time]
median_time
length(sample_A)
length(sample_B)
```

### Step 12: Perform Two-Sample K-S Test

Run the K-S test to check if `sample_A` and `sample_B` are drawn from the same continuous distribution.

```{r two-sample-ks-faithful}
ks_test_two_sample <- ks.test(sample_A, sample_B)
ks_test_two_sample
```

### Step 13: Interpret K-S Test 2

The Two-Sample Kolmogorov–Smirnov test was performed to check the Null Hypothesis (H_0) that the "shorter waiting times" (Sample A) and "longer waiting times" (Sample B) come from the same continuous distribution.

The result is highly significant:

- K-S Statistic (D): 1  
- P-value: \< 2.2e-16  

The test statistic D achieved its maximum value of 1, and the resulting p-value is effectively zero; we **strongly reject the Null Hypothesis**.

**Conclusion:** The distribution of the short waiting times and the distribution of the long waiting times are statistically distinct. This confirms that the `faithful` dataset is not just bimodal, but is likely a mixture of two separate (two different waiting time distributions operating within the same system).

### Step 14: Visualize the Two Separate Distributions (for pedagogical completeness)

```{r density-plot-faithful}
# 1. Create a data frame for Sample A and add a group label
df_A <- data.frame(
  waiting = sample_A,
  group = "Short Waits (<= 75 min)"
)

# 2. Create a data frame for Sample B and add a group label
df_B <- data.frame(
  waiting = sample_B,
  group = "Long Waits (> 75 min)"
)

# 3. Combine the two data frames
combined_samples_df <- rbind(df_A, df_B)

# Plot the densities on a single graph
ggplot(combined_samples_df, aes(x = waiting, fill = group)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Visual Confirmation of Two Distinct Distributions",
    x = "Waiting Time (minutes)",
    y = "Density",
    fill = "Sample Group"
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = c(
      "Short Waits (<= 75 min)" = "deepskyblue4",
      "Long Waits (> 75 min)"   = "darkorange3"
    )
  )
```

Why this is a Great Result?

- The graph clearly shows why we rejected the Null Hypothesis in both K-S tests.  
- We rejected the idea that the total data follows one Normal curve.  
- The image shows the two separate bell shapes that are actually present.  
- Two-Sample K-S Test (Step 12): We rejected the idea that the "Short Waits" (Blue) and "Long Waits" (Orange) come from the same distribution.  
- The graph clearly shows these two distributions are distinct and separate.  
- Confirms D = 1: The minimal overlap between the blue and orange curves, especially in the tails, is what drove your K-S statistic D to 1.
